On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-05-21 Manuscript/manuscript.tex
Removed commented out sections, fixed other issues

%Abstract should be 200 words
%Devon Allen was disqualified in 2022... discussions about dq, talk methods and 
%models, venue effect, and conclusion
\begin{abstract}
Devon Allen was disqualified in the men's 110 meter hurdle final of the 2022
World Track and Field Championships after registering a reaction time of 0.099 
seconds, 0.001 seconds faster than what is allowed. Following the games, 
bloggers on the running website LetsRun %scrutinized the results %and performed 
%non-statistical analysis of the data. They found 
concluded that the reaction time data from the 2022 World Championships 
seemed to be generally faster compared to the other datasets they considered, 
but they did not perform any formal %meaningful 
statistical analyses. This paper questions the reaction time 
disqualification barrier, which is currently 0.1 seconds, to determine whether 
this is a reasonable threshold. We employ a generalized linear mixed model 
(GLMM) with a random %effect term which will be referred to as the 
venue effect in order to %try and 
model the reaction time data. Additionally, we employ a signed-rank test for 
clustered data to 
compare reaction times for the same athletes at different competitions. This 
matter needs to be addressed because %Devon Allen's 
disqualification based on allowable reaction time 
%could be a 
%problem that could 
will continue to be an issue in future world championships.

%\eds{removed Devon Allen from keywords because his name is already in title
%and replaced with others}

%\eds{also removed 2022 World Track and Field Championships from keywords and 
%changed location of World in the title. World Track and Field Championships
%has more Google search results than Track and Field World Championships, so
%I opted to go with this. Also used plural Championships. Feel free to switch 
%order or use singular if one or both of these is 
%preferred for some reason.  }

@@

Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
%After coming in third in the 110 meter hurdles at the United States Track and 
%Field Championships also in Eugene at the end of June, Allen was poised to 
%compete in the 2022 World Track and Field Championships also 
%located at Eugene in August.
%two months later. 
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
%Allen, who attended the University of Oregon also in Eugene, 
%(the location of US World Championships 
%\eds{US World? remove World?} and the World Championship) 
%was
%considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
%in early June; only 0.04 seconds away from the world record \citep{wa2022preview}.
%Thus 
His disqualification in the final heat by reacting in 0.099 seconds instead of 
0.1 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification; it can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers 
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.

%2)
Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit, is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as \citet{johnson2022data}
and \citet{johnson2022was} from  Robert Johnson, who created and runs 
the LetsRun website.  He argued that there was an error
with the timing equipment at the 2022 World Track and Field Championships and
cited graphs, computed simple statistics such as median reaction times, and 
compared the reaction times at the 2022 World Track and Field Championships to 
other competitions
such as the United States Track and Field Championships and other World Championships.
The comparisons lacked formal statistical analysis, however. 
The alleged 99.95\% chance Allen was wronged 
came from Johnson claiming there is a $(1/13)^3$ chance that the fastest 
reaction times over the past 13 competitions of the men's 100 meter, men's 110 
meter hurdle, and women's 100 meter would all come from the same year 
\citep{johnson2022was}.
%came from Johnson claiming there is a $(1/13)^3$ chance that the fastest 
%reaction times came during the same championships for the men's 100 meter,
%men's 110 meter hurdle, and women's 100 meter for 13 competitions held between 2003 and 2022 
%\of{Cleaned up but may need additional work}
%\of{Cleaned this up a little but still a little messy.}
%\eds{is this better? (and still correct?)}
%\of{Yes, I think it looks good}

%3)
%The International Association of Athletics Federations (also referred to
%as IAAF or World Athletics), which is the governing board for international running
%competitions, commissioned a study in 2009 regarding the reaction time 
%barrier and whether the threshold should be lowered. 
%The paper analyzed sprinters reaction times and the biological processes 
%that went into reacting to the starting gun and concluded 
%that athletes can react in as low as 0.08 seconds.  
%One study by Pain looked at
Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined
reaction times for nine male athletes.  The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by International Association of Athletics Federations 
(IAAF), \citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
\eds{are hurdlers generally considered a subset of sprinters or are they 
considered a separate type of athlete?  Specifically did the study above 
recommend the barrier be lowered for both sprinters and hurdlers?
You also refer to sprinters elsewhere in the manuscript.  Are those instances
meant to refer to both sprinters and hurdlers, as well?}
\of{Yes I would say that hurdlers are a subset of sprinter.  The reaction time
barrier for the men's 100 meters and men's 110 meter hurdle are both 0.1 seconds.
The paper recommended for the reaction time barrier to be decreased to roughly 0.8
seconds.  It does not reference any distinction between hurdlers/sprinters, so I
think they can be grouped together.  In my personal opinion I don't see why the
reaction time for hurdlers versus 100m dash would differ significantly.  In terms
of what I wrote in the paper, I want to stay accurate to what is in the study
but also generalize studies involving reaction times to include hurdlers.  I believe
many of the studies examining the reaction time barrier from a physiological
perspective to find the minimum possible reaction time do not deal with hurdles.
Thus, I don't know if I can change what I wrote to be both accurate and understandable,
but what do you both think?}
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

@@

%This will
%be important when considering the results in section \ref{sec:Results}.

Despite their study, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.  It is unclear
why IAAF would commission a study only to disregard the results and not
make the recommended changes.  It is unknown how many athletes since 2009 could
have potentially benefited from this, but it is probable that there are others
besides Allen who were disqualified.
%Reaction time barrier has been examined before: cite reviews
%Then go into specifics and talk about people who wanted to change barrier
%General reviews are important
%People have studied the biological processes, cite a few references
%How barrier is picked by regulator
%Read abstract and into and summarize into about one sentence what each paper contributes
%1 and 2 can be merged but 3 need to stand alone

%4)
The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championships to check for any abnormalities, and we approach
this in two different ways.  First we model the data, which are
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with a random effect 
accounting for the venue (or year). 
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Second, we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing reaction times from the same athlete across multiple competitions we 
are able to determine if the 2022 reaction times differed significantly from 
those of other competitions.
Our investigation into reaction times is important not only because of its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times, which is the goal of
every sprinter who steps onto the track.

@@

Data was taken from the World Athletics 
%\eds{is there a website you can cite here?}
and covers the men's 110 meter hurdles from 1999 to 2022; the source can be 
found in the Appendix. %~\ref{sec:Appendix}.   %\eds{see earlier comment - 
%are you using any of these as covariates?}
%\of{removed covariate confusion, talk about womens a couple sentences down} 
Although women's data is available, we chose not to include women's data in the
analyses to follow due to the numerous papers citing differences in men and
women's reaction times \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender}.
%The variable ``Stage" refers to whether the observation occurred during a 
%heat (preliminary round)  
%semifinal heat (of which there are either two
%or three each year) denoted by ``S", or final heat denoted by ``F".
We focus on the reaction times during the semifinal and final heats only, as 
reaction times from preliminary heats are often not as fast as those in later 
heats %were not studied because athletes react faster as the 
%competition progresses 
\citep[e.g.,][]{zhang2021correlation}. 
%The variable 
%``Gender" has levels ``M" and ``F" for ``male" and ``female" respectively.  
%The variable ``Batch" takes a numerical value and every heat of every round from 
%1999 to 2022 was assigned it's
%own number.  This was added because of suspected issues that reaction time may be
%correlated with other runners reacting quickly (one runner reacting quickly may
%result in other runners reacting quickly). There is no statistical meaning to
%how the batches were labeled, they were ordered chronologically by meet but
%reverse chronologically by year (Batch 1 is a preliminary heat of 2022 and Batch
%111 corresponds to the 1999 finals). %Included are
%box plots of the reaction times of the men's finals and the men's semifinals and
%finals pooled together.  
For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together and refer to this data as the ``pooled data"
%We will refer to the latter as the ``pooled data"
throughout the paper.  %The heats or preliminary rounds were not studied because
 % Athletes react 
%faster in the finals than the semifinals which are faster than the heats 
%\citep{zhang2021correlation}.  
%\eds{if you are really not using ``H'' data at all, it does not appear necessary
%to spend time defining earlier.  You could just say directly when you define 
%stage that you only consider semifinals and finals and cite the Zhang paper for 
%rationale}
%\of{Ok I updated this.  Should I delete the batch variable too?  Although I don't
%use it, I do reference it in the methods section}
The pooled data increases the sample size which is advantageous because in some
years there are very few final heat observations.  For example, in 2022 there
were only five data points due to two disqualifications and one athlete who 
did not compete.

@@

\eds{I wonder if this paragraph would be better placed in the introduction?}
\of{I think it could fit in the brief section between the start of the Data section
and the GLMM subsection in the paragraph starting with "There are two types..."}
\eds{With the following newly added sentences, I think we can leave this paragraph here.}
\of{I agree I think this works well.}
We can also see from Figure~\ref{fig:Boxplots} that the reaction times clearly
differ from year to year. The venue of the World Championships changes each 
year, so weather or climate related factors such as humidity, precipitation, 
elevation, may indeed play a role, but additionally, both the technology and 
false start penalization rules have changed over the study period. 
%Something that is extremely important to note in the data is the inconsistency 
%of World Athletics in terms of how they penalize false starts.  
From 2007--2009,
the IAAF (the former name for World Athletics, and the governing body for the 
World Track and Field Championships) instituted a rule change that allowed one 
warning false start before a sprinter was disqualified \citep{iaaf2009falsestart}. 
%This was desirable for World Athletics as false starts can make already long 
%track meets more tedious, both for athletes and viewers watching the television 
%broadcast.  
%\eds{I think we lost something here with all of the edits so I removed the 
%previous sentence. What was desirable for World Athletics -- the 2007-2009 rule
% change?  Why was this desirable?  The rule change would seem to make the event
% longer and not shorter? } 
There were 18 male and 7 female false starts at the 2007 World 
Championships, and 18 male and 7 female false starts at the 2009 World 
Championships. 
Starting in 2011, the rule was scrapped, and the old policy which automatically 
disqualifies runners who false start was reinstated. By returning to the harsher 
policy and cracking down on false starts, World Athletics had reduced men's 
false starts by two thirds in 2011 (6 male and 4 female disqualifications) 
\citep{iaaf2009falsestart}.
%The effect? At the 2011 
%World Championships there were 6 male disqualifications and 4 female 
%disqualifications \citep{iaaf2009falsestart}.  
%One study that looked at different false start rules that IAAF has
%imposed from 1997 to 2009 and found that the reaction times when there was less 
%leniency improved by a statistically significant amount
%\citep{haugen2013effect}. 
\citet{haugen2013effect} examined the effect of different false start
rules that IAAF has imposed from 1997 to 2009 and found statistically 
significant improvement in reaction times under more lenient rules.

\subsection{Beyond World Championships}\label{sec:databeyond}
The motivation for looking beyond World Championships data came from an 
exploratory analysis examining how United States athletes reacted at the 
2022 United States Track and Field Championships.
%\eds{be consistent with how you refer to the US championship:
%``United States Track and Field Championships" and ``United States Track and Field 
%Championships" are both being used; please pick one and use that throughout}
%\of{Resolved; changed all USA to United States.}
From June 23-26 2022, the United States held its Track
and Field Championships at Hayward 
Field in Eugene, Oregon to decide who to send to the World Championships 
being held at the same venue in August. %the exact same venue as used 
%for the World Track and Field Championships a month later.  
%Thus, a baseline is able to be established for the four United States athletes who 
Thus, we can establish a baseline for the four United States athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championships races compared to the United States Track and Field 
Championships races. Devon Allen raced three times at the United States Track and
Field Championships and three times at the World Track and Field Championships. 
His reaction times were substantially faster in the World Track and
Field Championships compared to the United States Track and Field Championships, 
as were his
teammates. 
This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  Table~\ref{fig:USAvsWorld} highlights the difference in reaction 
times for Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts.

@@

%\of{Im not sure how to fix this but the spacing between the lines of the
%table is different than the rest of the paper}
%\eds{LaTeX keeps captions singlespaced automatically with the doublespace 
%command; don't worry about it!}

\eds{this is where you can put the relevant 
information about the other competitions that was commented out in the 
introduction.}
\eds{Can you show some descriptive statistics for this data?  Total number of 
clusters? Average number of international vs national competitions across 
athletes/clusters?}
\of{Is the one below what you're talking about?}
\eds{Yes, this is the part from the intro I was referring to, but you need to 
edit this paragraph now so it all works together.}
The dataset of the four United States athletes 
is too small to perform any meaningful statistical analysis, but
the data was expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete). First we can compare the reaction times of athletes who
competed at the World Track and Field Championships in both 2019 and 2022, and 
then we can compare how athletes performed at the 2022 championships relative 
to their national/qualifiers meet. In order to reduce any chances
of athletes improving reaction time, we can also look at data from qualifying meets.
Almost every athlete had to qualify by proving they are elite at the national level, 
usually by coming in at least fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August. The equipment used to measure reaction
time cannot be guaranteed to be the same from national to international
competition, but the time span between meets has been reduced considerably.

@@

The %World Championships data and the data eyond the World Championships 
data described in Sections~\ref{sec:dataworld} and~\ref{sec:databeyond} were
analysed with GLMM and rank-based methods, respectively.
%\eds{need consistency with ``Championship" or ``Championships - I am not sure 
%which one is correct.  I am using Championships (plural) but
%change everything to singular if that is preferred}
%\of{I believe that championships is correct, I have changed all instances of
%championship to championships}

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. A random effect for the venue or year was also found to be necessary, 
which makes intuitive sense given the discussion in Section~\ref{sec:dataworld}. 
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
\eds{we know that many of the athletes compete in multiple world championships,
  how is this athlete clustering being accounted for here?}
\jy{We could add an athlete random effect too, if the players are identified in the data.}
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by 
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ij}$ has conditional mean
\[
g\{E(Y_{ij} | z_i)\} = \alpha + z_i,
\]
and dispersion parameter $\phi$, where $g$ is a known link function and
$z_i$ is a normally distributed random effect with mean zero and
variance~$\sigma^2$. Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time data, the logarithm 
link function worked better for our data and additionally ensures the positivity
of the conditional mean.
% The link function $g$ can be the logarithm function to
% ensure the positivity of the conditional mean.
%\eds{Did you try an identity link also? Lo and Andrews (2015) suggest that the 
%Gamma family with an identity link works well for their reaction-time data.
%It doesn't seem like a good idea in general because it is likely that some 
%negative predicted responses will occur either during model fitting or in the 
%final model, but the internet seems to favor this link so a reviewer may ask 
%about it too.} \of{I have added code at the bottom of Code/Models.Rmd that
%compared the link = "identity" and link="log" for each dataset.  I found that
%the link=log was marginally better.}
%We include a random effect for venue (or year) because the technology
%that World Athletics has used has differed since 1999, and theoretically the
%reaction time data from 2022 should be the best as the timing system should be
%state of the art. Additionally, any other weather or climate related factors 
%such as humidity, precipitation, elevation, could be captured by this effect.
%\eds{I moved this info into the Data section.}
The gamma mixed-effect model can be fit with the \texttt{glmer()} function in R
package \texttt{lme4} \citep{lme4}.

@@

Finally, we remark here that it is also possible to include a heat or race 
\eds{heat? when I first read this, I thought you meant race as in ethnicity}
\of{Yes I think heat or batch effect works better, especially now that I have
removed most talk about heat from the data section} 
\eds{I see, I did not understand this to be Batch effect when I read it the 
first time.  I commented out the Batch text in the Data section - I think it is 
fine to just discuss it here, as it ultimately was not used.  I will defer to 
Dr Yan on this, however.}
effect to account for some heats possibly having faster or slower 
reaction times for everyone involved in the heat.
%From our exploration, however, the race effect is very difficult to explain or 
%interpret as it means that some heats of races were either
%faster or slower and that everyone in the race reacted similarly.  
Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other. 
\eds{Did you try it and find that the effect was not needed?  If yes,
I would include that here.  That argument 
would be more compelling for its non-use than its difficulty in interpretation.}

@@

%% In this case, athletes who have more
%% data are generally thought to be faster, as they have made it further in the
%% competition.  Clusters range from two to six, with two indicating that at each
%% competition they only made it to the heats, but a cluster size of six indicates
%% that they made the finals in both competitions.  While the number of times
%% athletes compete is based off their overall time not their reaction time, the
%% two are correlated \citep{delalija2008reaction}.
%% Group is either designated to be 1 or 2, with 1 corresponding to either the
%% national level data or the previous world championships data and 2 corresponding
%% to the 2022 data.

>>>>

%Abstract should be 200 words
%Devon Allen was disqualified in 2022... discussions about dq, talk methods and 
%models, venue effect, and conclusion
\begin{abstract}
Devon Allen was disqualified in the men's 110 meter hurdle final of the 2022
World Track and Field Championships after registering a reaction time of 0.099 
seconds, 0.001 seconds faster than what is allowed. Following the games, 
bloggers on the running website LetsRun concluded that the reaction time data 
from the 2022 World Championships seemed to be generally faster compared to the 
other datasets they considered, but they did not perform any formal 
statistical analyses. This paper questions the reaction time 
disqualification barrier, which is currently 0.1 seconds, to determine whether 
this is a reasonable threshold. We employ a generalized linear mixed model 
(GLMM) with a random venue effect in order to  model the reaction time data. 
Additionally, we employ a signed-rank test for clustered data to 
compare reaction times for the same athletes at different competitions. This 
matter needs to be addressed because disqualification based on allowable 
reaction time will continue to be an issue in future world championships.

@@

Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
His disqualification in the final heat by reacting in 0.099 seconds instead of 
0.1 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification; it can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers 
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.

%2)
Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit, is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as \citet{johnson2022data}
and \citet{johnson2022was} from  Robert Johnson, who created and runs 
the LetsRun website.  He argued that there was an error
with the timing equipment at the 2022 World Track and Field Championships and
cited graphs, computed simple statistics such as median reaction times, and 
compared the reaction times at the 2022 World Track and Field Championships to 
other competitions
such as the United States Track and Field Championships and other World Championships.
The comparisons lacked formal statistical analysis, however. 
The alleged 99.95\% chance Allen was wronged 
came from Johnson claiming there is a $(1/13)^3$ chance that the fastest 
reaction times over the past 13 competitions of the men's 100 meter, men's 110 
meter hurdle, and women's 100 meter would all come from the same year 
\citep{johnson2022was}.

Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined
reaction times for nine male athletes.  The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by International Association of Athletics Federations 
(IAAF), \citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
\eds{are hurdlers generally considered a subset of sprinters or are they 
considered a separate type of athlete?  Specifically did the study above 
recommend the barrier be lowered for both sprinters and hurdlers?
You also refer to sprinters elsewhere in the manuscript.  Are those instances
meant to refer to both sprinters and hurdlers, as well?}
\of{Yes I would say that hurdlers are a subset of sprinter.  The reaction time
barrier for the men's 100 meters and men's 110 meter hurdle are both 0.1 seconds.
The paper recommended for the reaction time barrier to be decreased to roughly 0.8
seconds.  It does not reference any distinction between hurdlers/sprinters, so I
think they can be grouped together.  In my personal opinion I don't see why the
reaction time for hurdlers versus 100m dash would differ significantly.  In terms
of what I wrote in the paper, I want to stay accurate to what is in the study
but also generalize studies involving reaction times to include hurdlers.  I believe
many of the studies examining the reaction time barrier from a physiological
perspective to find the minimum possible reaction time do not deal with hurdles.
Thus, I don't know if I can change what I wrote to be both accurate and understandable,
but what do you both think?}
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

@@

Despite their study, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.  It is unclear
why IAAF would commission a study only to disregard the results and not
make the recommended changes.  It is unknown how many athletes since 2009 could
have potentially benefited from this, but it is probable that there are others
besides Allen who were disqualified.

The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championships to check for any abnormalities, and we approach
this in two different ways.  First we model the data, which are
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with a random effect 
accounting for the venue (or year). 
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Second, we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing reaction times from the same athlete across multiple competitions we 
are able to determine if the 2022 reaction times differed significantly from 
those of other competitions.
Our investigation into reaction times is important not only because of its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times, which is the goal of
every sprinter who steps onto the track.

@@

Data was taken from the World Athletics 
%\eds{is there a website you can cite here?}
and covers the men's 110 meter hurdles from 1999 to 2022; the source can be 
found in the Appendix.
Although women's data is available, we chose not to include women's data in the
analyses to follow due to the numerous papers citing differences in men and
women's reaction times \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender}.
We focus on the reaction times during the semifinal and final heats only, as 
reaction times from preliminary heats are often not as fast as those in later 
heats \citep[e.g.,][]{zhang2021correlation}.

For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together and refer to this data as the ``pooled data"
throughout the paper. 
%\of{Ok I updated this.  Should I delete the batch variable too?  Although I don't
%use it, I do reference it in the methods section}
The pooled data increases the sample size which is advantageous because in some
years there are very few final heat observations.  For example, in 2022 there
were only five data points due to two disqualifications and one athlete who 
did not compete.

@@

\eds{I wonder if this paragraph would be better placed in the introduction?}
\of{I think it could fit in the brief section between the start of the Data section
and the GLMM subsection in the paragraph starting with "There are two types..."}
\eds{With the following newly added sentences, I think we can leave this paragraph here.}
\of{I agree I think this works well.}
We can also see from Figure~\ref{fig:Boxplots} that the reaction times clearly
differ from year to year. The venue of the World Championships changes each 
year, so weather or climate related factors such as humidity, precipitation, 
elevation, may indeed play a role, but additionally, both the technology and 
false start penalization rules have changed over the study period.

From 2007--2009, the IAAF (the former name for World Athletics, and the 
governing body for the World Track and Field Championships) instituted a rule 
change that allowed one warning false start before a sprinter was disqualified 
\citep{iaaf2009falsestart}. There were 18 male and 7 female false starts at the 
2007 World Championships, and 18 male and 7 female false starts at the 2009 World 
Championships. Starting in 2011, the rule was scrapped, and the old policy which
automatically disqualifies runners who false start was reinstated. This was 
desirable for World Athletics as false starts can make already lengthy track 
meets more tedious, both for athletes and viewers watching the television 
broadcast. By returning to the harsher policy and cracking down on false starts,
World Athletics had reduced men's false starts by two thirds in 2011 (6 male and
4 female disqualifications) \citep{iaaf2009falsestart}. \citet{haugen2013effect}
examined the effect of different false start rules that IAAF has imposed from 
1997 to 2009 and found statistically significant improvement in reaction times 
under more lenient rules.

\subsection{Beyond World Championships}\label{sec:databeyond}
The motivation for looking beyond World Championships data came from an 
exploratory analysis examining how United States athletes reacted at the 
2022 United States Track and Field Championships.

From June 23-26 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World Championships 
being held at the same venue in August. Thus, we can establish a baseline for the four United States athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championships races compared to the United States Track and Field 
Championships races. Devon Allen raced three times at the United States Track and
Field Championships and three times at the World Track and Field Championships. 
His reaction times were substantially faster in the World Track and
Field Championships compared to the United States Track and Field Championships, 
as were his
teammates. 
This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  Table~\ref{fig:USAvsWorld} highlights the difference in reaction 
times for Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts.

@@

\eds{this is where you can put the relevant 
information about the other competitions that was commented out in the 
introduction.}
\eds{Can you show some descriptive statistics for this data?  Total number of 
clusters? Average number of international vs national competitions across 
athletes/clusters?}
\of{Is the one below what you're talking about?}
\eds{Yes, this is the part from the intro I was referring to, but you need to 
edit this paragraph now so it all works together.}

\begin{figure}[tbp]
  \centering
  \includegraphics{2019vs2022BarGraph}
  \caption{The reaction times from 2019 and 2022}
  \label{fig:2019vs2022Graph}
\end{figure}

The dataset of the four United States athletes 
is too small to perform any meaningful statistical analysis, but
the data was expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete). First we can compare the reaction times of athletes who
competed at the World Track and Field Championships in both 2019 and 2022, and 
then we can compare how athletes performed at the 2022 championships relative 
to their national/qualifiers meet. In order to reduce any chances
of athletes improving reaction time, we can also look at data from qualifying meets.
Almost every athlete had to qualify by proving they are elite at the national level, 
usually by coming in at least fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August. The equipment used to measure reaction
time cannot be guaranteed to be the same from national to international
competition, but the time span between meets has been reduced considerably.

@@

The data described in Sections~\ref{sec:dataworld} and~\ref{sec:databeyond} were
analyzed with GLMM and rank-based methods, respectively.

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. A random effect for the venue or year was also found to be necessary, 
which makes intuitive sense given the discussion in Section~\ref{sec:dataworld}. 
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
\eds{we know that many of the athletes compete in multiple world championships,
  how is this athlete clustering being accounted for here?}
\jy{We could add an athlete random effect too, if the players are identified in the data.}
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by 
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ij}$ has conditional mean
\[
g\{E(Y_{ij} | z_i)\} = \alpha + z_i,
\]
and dispersion parameter $\phi$, where $g$ is a known link function and
$z_i$ is a normally distributed random effect with mean zero and
variance~$\sigma^2$. Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time data, the logarithm 
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

@@

Finally, we remark here that it is also possible to include a heat or race 
\eds{heat? when I first read this, I thought you meant race as in ethnicity}
\of{Yes I think heat or batch effect works better, especially now that I have
removed most talk about heat from the data section} 
\eds{I see, I did not understand this to be Batch effect when I read it the 
first time.  I commented out the Batch text in the Data section - I think it is 
fine to just discuss it here, as it ultimately was not used.  I will defer to 
Dr Yan on this, however.}
effect to account for some heats possibly having faster or slower 
reaction times for everyone involved in the heat.

Although it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other. 
\eds{Did you try it and find that the effect was not needed?  If yes,
I would include that here.  That argument 
would be more compelling for its non-use than its difficulty in interpretation.}

@@
